---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# KOMAP

<!-- badges: start -->
<!-- badges: end -->

The goal of KOMAP is to fit an unsupervised phenotyping system which does not require manual chart review to acquire gold standard labels for algorithm training and validation. The online algorithm only requires summary-level covariance matrix for coefficient training. To get a quick model evaluation without individual-level data, users can upload conditional mean vector and covariance matrix to obtain a simulated model AUC. KOMAP can deal with cases when both codified and NLP features are available, or when only one feature type is at hand (just leave `target.code` or `target.cui` equal to `NULL` if not available). 

## Installation

You can install the development version of KOMAP from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("xinxiong0238/KOMAP")
```


## Data preprocessing

### Input covariance matrix (wide format)
One of the KOMAP algorithm inputs is the covariance matrix of the clinical features (i.e., log(count+1)) derived from your target population. To obtain the covariance matrix, we start from the raw EHR records. Note in the stored fake ehr data, we only include 5 patients and 2 different ehr codes - lab code and diagnostic ICD code (DIAG-ICD9).
```{r}
library(KOMAP)
head(fake_ehr)
```

We need to first calculate the raw concept count once per day for each patient(i.e., for those raw codes that show up more than one time in a day, treat as one time):

```{r message=FALSE, warning=FALSE}
library(dplyr)
fake_ehr_raw_count <- fake_ehr %>% 
  group_by(patient_num, concept_type, concept_code) %>%
  summarise(count_once_per_day = length(unique(days_since_admission)))
head(fake_ehr_raw_count)
```

Then for diagnostic ICD code, we need to roll them up to phecode level. An toy example of the mapping file is stored in the package:
```{r}
head(phecode_map)
```

For each patient, the phecode count is equal to the sum of all its descendant ICD code once-per-day count:

```{r message=FALSE, warning=FALSE, include=FALSE}
fake_ehr_rollup_count_phecode <- left_join(fake_ehr_raw_count %>% 
                                            filter(concept_type == 'DIAG-ICD9'), 
                                           phecode_map, multiple = "all")
fake_ehr_rollup_count_phecode <- fake_ehr_rollup_count_phecode %>%
  group_by(patient_num, concept_type, Phecode) %>%
  summarise(count = sum(count_once_per_day)) %>%
  mutate(concept_code = Phecode) %>%
  select(!Phecode)
```

Then, we combine the rolled up phecode count and raw loinc count together:
```{r}
fake_ehr_count_loinc <- fake_ehr_raw_count %>% filter(concept_type == 'LAB-LOINC') %>%
  mutate(count = count_once_per_day,
         concept_code = paste0('LOINC:', concept_code)) %>%
  select(!count_once_per_day)
fake_ehr_count <- rbind(fake_ehr_count_loinc, fake_ehr_rollup_count_phecode)
head(fake_ehr_count)
```

Then we transform the data to the wide format, conduct logarithm transformation:
```{r}
fake_ehr_count_wide <- tidyr::pivot_wider(fake_ehr_count, id_cols = c('patient_num'),
                                         names_from = 'concept_code', values_from = 'count')
fake_ehr_count_wide[is.na(fake_ehr_count_wide)] <- 0
fake_ehr_logcount_wide <- fake_ehr_count_wide
fake_ehr_logcount_wide[,-1] <- log(fake_ehr_logcount_wide[, -1] + 1)
```

Note that it may be beneficial to screen out patients who never have the diagnostic code of your target disease before calculating the covariance matrix. For example, if you are interesting in rheumatoid arthritis phenotyping task, it is suggested to focus on patients who had at least 1 count of `PheCode:714.1` in their EHR records.

#### Corrupt main surrogate
In order to fine tune the parameter from summary-level statistics, we need to construct corrupted surrogate features before calculating the covariance matrix. Specifically, for each main surrogate feature, we set 20% of data to be equal to the mean of the feature as if performing dropout training. Then we further split up the data into two parts (1/3 and 2/3) to construct training and validation covariance matrices:

```{r}
main_code = 'PheCode:250'
fake_ehr_corrupt = as.data.frame(fake_ehr_logcount_wide)
fake_ehr_corrupt$corrupt_mainICD = fake_ehr_corrupt[, main_code]
fake_ehr_corrupt$corrupt_mainICD[sample(1:nrow(fake_ehr_corrupt), round(nrow(fake_ehr_corrupt) * 0.2), replace = FALSE)] = mean(as.matrix(fake_ehr_corrupt[, main_code]))

id.train = sample(1:nrow(fake_ehr_corrupt), round(nrow(fake_ehr_corrupt) / 3 * 2))
id.valid = setdiff(1:nrow(fake_ehr_corrupt), id.train)
dat.cov.train = cov(fake_ehr_corrupt[id.train, ])
dat.cov.valid = cov(fake_ehr_corrupt[id.valid, ])
```


### One convenient function
To wrap up the aforementioned preprocessing steps, the KOMAP package provides one function `gen_cov_input` with a raw longitudinal ehr data, one rollup dictionary and some frequency filters as the function input:
```{r message=FALSE}
data(ehr_data)
data(rollup_dict)
data(filter_df)
input_cov <- gen_cov_input(ehr_data, rollup_dict, filter_df, main_surrogates = 'PheCode:250', train_ratio = 1/2)
input_cov$train_cov[1:3,1:3]
input_cov$valid_cov[1:3,1:3]
```



### Input covariance matrix (long format)
You can also input the covariance matrix in a long format. The data must contain three columns with the first two indicating the node pair names. Note that when KOMAP tranforms the input to a wide format, it will assign 0 to missing pairwise covariance value. If you only input the lower/upper triangle part, the function will automatically fill the other side assuming symmetric covariance matrix. See an example below:

```{r}
head(cov_RA_train_long)
```



### Input feature filters (optional)
To further filter out unrelated codes, you can specify a vector of codified features as well as a vector of NLP features (e.g., CUIs in the stored example). You can refer to [ONCE](https://shiny.parse-health.org/ONCE/), an online searching app that returns a dictionary of related codified features as well as NLP features based on knowledge extracted from online article corpus. The csv file you downloaded from ONCE app has the same format as `codify_RA` (or `cui_RA`) stored in the KOMAP pacakge.
```{r}
codify.feature <- codify_RA$Variable[codify_RA$high_confidence_level == 1]
head(codify.feature)
nlp.feature <- cui_RA$cui[cui_RA$high_confidence_level == 1]
head(nlp.feature)
```


If `codify.feature` or `nlp.feature` argument is not equal to NULL, then KOMAP will only take a subset of the covariance matrix corresponding to your input. You can also do the filtering on the input covariance matrix by yourself.


### Input conditional covariance matrix, conditional mean vector and prevalance (optional)
If you have some labeled data at hand and you want to quickly check KOMAP performance, this package provides a simulated AUC that only needs conditional covariance matrix, conditional mean vector and prevalence without any individual data. To generate such AUC, we start from `fake_ehr_label_logcount_wide` which was generated by the aforementioned process using another set of fake patients. The format is very similar to `fake_ehr_logcount_wide`, except that `fake_ehr_label_logcount_wide` has an additional column indicating the true label for each patient.
```{r}
fake_ehr_label_logcount_wide[1:3,1:5]
```

The conditional statistics are derived based on this label column.

```{r}
fake_ehr_var0 = cov(fake_ehr_label_logcount_wide %>% filter(Y == 0) %>% select(-c(Y, patient_num)))
fake_ehr_var1 = cov(fake_ehr_label_logcount_wide %>% filter(Y == 1) %>% select(-c(Y, patient_num)))
fake_ehr_mu0 = colMeans(fake_ehr_label_logcount_wide %>% filter(Y == 0) %>% select(-c(Y, patient_num)))
fake_ehr_mu1 = colMeans(fake_ehr_label_logcount_wide %>% filter(Y == 1) %>% select(-c(Y, patient_num)))
fake_ehr_prev = mean(fake_ehr_label_logcount_wide$Y)
```

Note that the two conditional matrices must have colnames and rownames, and the conditional mean vectors must have names.


### Input gold label individual data (optional)
If you want to obtain the accurate AUC value by providing individual count data, the input gold label data should have the format as `fake_ehr_label_logcount_wide` with one column indicating patient labels (the column name should go to the `nm.y` argument). Additionally, you can have another column suggesting the sample probability (the column name should go to the `nm.pi` argument) if each patient in the label data was sampled with unequal probability.

```{r}
fake_ehr_label_logcount_wide_pi[1:3,1:6]
```

### Input individual data for disease score prediction (optional)
If you have some unlabeled data and want to use KOMAP to perform disease score prediction, the input data should be similar to `fake_ehr_logcount_wide`. If only part of your individual data has labels (i.e., some patients have NA value on the `Y` column), KOMAP can still perform prediction by ignoring the `Y` column. You also need to assign the unique patient id to a column (i.e, `nm.id` argument) for your input individual data, so that the output prediction data frame will also have that column in case of mismatching issue.



## Toy example for how use stored input data

### Stored input data for phenotyping
To illustrate the requirement for input data format, users can refer to the toy examples stored in the package.

```{r}
library(KOMAP)
## Input covariance matrix
?cov_RA

## Input dictionary (can be NULL)
?dict_RA

## Input individual data to predict disease score and/or evaluate model performance (can be NULL)
?dat_part

```

## Phenotyping Example

This is a basic example of how to use the main function in the KOMAP package to perform model training, score prediction and model evaluation with built-in toy rheumatoid arthritis data.

### Only output with regression coefficients
```{r example}
codify.feature <- codify_RA$Variable[codify_RA$high_confidence_level == 1]
nlp.feature <- cui_RA$cui[cui_RA$high_confidence_level == 1]
input.cov.train <- cov_RA_train_long
input.cov.valid <- cov_RA_valid_long

target.code <- 'PheCode:714.1'
target.cui <- 'C0003873'
nm.corrupt.code <- 'corrupt_mainICD'
nm.corrupt.cui <- 'corrupt_mainNLP'
nm.utl <- 'utl'
nm.pi <- 'pi'
nm.id <- 'patient_num'
nm.y <- 'Y'
dat.part <- dat_part

## When the input is in a long format:
out_input_long <- KOMAP_corrupt(input.cov.train, input.cov.valid, is.wide = FALSE, target.code, target.cui, 
                                nm.disease = 'RA', nm.utl, nm.multi = NULL, nm.corrupt.code = nm.corrupt.code, 
                                nm.corrupt.cui = nm.corrupt.cui, dict_RA, pred = FALSE, 
                                eval.real = FALSE, eval.sim = FALSE)
```
If you do not specify `pred`, `eval.real` or `eval.sim` equal to be `TRUE`, then KOMAP only output the regression coefficients which can be used for disease scoring calculation with additional individual data. Inside the `est` output, the `long_df` data frame in the output pulls coefficients from different model settings together if any (e.g., only use main ICD as surrogate or use both main ICD and main NLP as surrogates), while `lst` organizes the regression output by model settings.

```{r}
head(out_input_long$est$long_df)
```

Similarly, you can input the covariance matrix in wide format:
```{r}
input.cov.train <- cov_RA_train
input.cov.valid <- cov_RA_valid
out_0 <- KOMAP_corrupt(input.cov.train, input.cov.valid, is.wide = TRUE, target.code, target.cui, 
                nm.disease = 'RA', nm.utl, nm.multi = NULL, nm.corrupt.code = nm.corrupt.code, 
                nm.corrupt.cui = nm.corrupt.cui, dict_RA, pred = FALSE, 
                eval.real = FALSE, eval.sim = FALSE)
```



You can also specify a list of codified/narrative features that are relevant to the target disease. KOMAP will perform feature screening before regression:
```{r}
out_1 <- KOMAP_corrupt(input.cov.train, input.cov.valid, is.wide = TRUE, target.code, target.cui, 
                       nm.disease = 'RA', nm.utl, nm.multi = NULL, nm.corrupt.code = nm.corrupt.code, 
                       nm.corrupt.cui = nm.corrupt.cui, dict_RA, 
                       codify.feature = codify.feature, nlp.feature = nlp.feature,
                       pred = FALSE, eval.real = FALSE, eval.sim = FALSE)
```


### Output regression coefficients and simulated AUC

When specifying `eval.sim=TRUE` and input conditional mean vectors, conditional covaraince matrices and the prevalence value, KOMAP will return not only the regression coefficients, but also a simulated AUC that helps a grasp of model performance without any individual labeled data.
```{r}
out_2 <- KOMAP_corrupt(input.cov.train, input.cov.valid, is.wide = TRUE, target.code, target.cui, 
                       nm.disease = 'RA', nm.utl, nm.multi = NULL, nm.corrupt.code = nm.corrupt.code, 
                       nm.corrupt.cui = nm.corrupt.cui, dict_RA, 
                       codify.feature = codify.feature, nlp.feature = nlp.feature,
                       pred = FALSE, eval.real = FALSE, eval.sim = TRUE,
                       mu0, mu1, var0, var1, prev_Y, B = 10000)
out_2$sim_eval
```

### Output regression coefficients and predicted disease scores
When specifying `pred=TRUE` and input a group of individual data (i.e., log count of EHR features), KOMAP will return both the regression coefficients, predicted disease scores and predicted disease labels (using gaussian mixture model) for the input patients. Remember to specify the `nm.id` argument (unique patient id) to avoid score matching issue.
```{r}
library(mclust)
out_3 <- KOMAP_corrupt(input.cov.train, input.cov.valid, is.wide = TRUE, target.code, target.cui, 
                       nm.disease = 'RA', nm.utl, nm.multi = NULL, nm.corrupt.code = nm.corrupt.code, 
                       nm.corrupt.cui = nm.corrupt.cui, dict_RA, 
                       codify.feature = codify.feature, nlp.feature = nlp.feature,
                       pred = TRUE, eval.real = FALSE, eval.sim = FALSE,
                       dat.part = dat.part, nm.id = nm.id)
head(out_3$pred_prob$pred.score)
head(out_3$pred_prob$pred.cluster)
```

### Output regression coefficients and true AUC
When specifying `eval.real=TRUE` and input a group of individual data (i.e., log count of EHR features) as well as their labels, KOMAP will return both the regression coefficients, and real AUC evaluated based on your gold label data. Remember to specify the `nm.y` argument (gold label for each patient). It will also aumatically return the predicted disease score for each patient.
```{r}
out_4 <- KOMAP_corrupt(input.cov.train, input.cov.valid, is.wide = TRUE, target.code, target.cui, 
                       nm.disease = 'RA', nm.utl, nm.multi = NULL, nm.corrupt.code = nm.corrupt.code, 
                       nm.corrupt.cui = nm.corrupt.cui, dict_RA, 
                       codify.feature = codify.feature, nlp.feature = nlp.feature,
                       pred = FALSE, eval.real = TRUE, eval.sim = FALSE,
                       dat.part = dat.part, nm.id = nm.id, nm.pi = nm.pi, nm.y = nm.y)
out_4$real_eval
```
